Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-02-26 Nicolas.tex
1 column acceptance

\abstract{The detailed nature of type Ia supernovae (SNe~Ia) remains uncertain,
    and as survey statistics increase, the question of astrophysical systematic
    uncertainties rises, notably that of SN~Ia population evolution. In this
    paper, we study the dependence with redshift of the SN~Ia \texttt{SALT2.4}
    lightcurve stretch, a purely intrinsic SN property, to probe its potential
    redshift drift. The SN stretch has been shown to strongly correlate with the
    SN environment, notably with stellar age tracers. We model the underlying
    stretch distribution as a function of redshift, using the evolution of the
    fraction of young and old SNe~Ia as predicted \textbf{in a previous study},
    and assuming constant underlying stretch distribution for each age
    population consisting of Gaussian mixtures. We test our prediction against
    published samples cut to have marginal magnitude selection effects, so that
    any observed change is indeed of astrophysical and not observational in
    origin. In this first study, there are indications that the underlying
    SN~Ia stretch distribution is evolving as a function of redshift, and that
    the young/old drifting model is a better description of the data than any
    time-constant model, including the sample-based asymmetric distributions
    often used to correct Malmquist bias at a significance of over 5 $\sigma$.
    The favored underlying stretch model is \textbf{a bimodal one}, composed of
    a high-stretch mode shared by both young and old environments, and a
    low-stretch mode exclusive to old environments. The precise impact of the
    redshift evolution of the SN Ia population intrinsic properties on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
    distances derived from SN affected by observational selection effects. We
highlight that such a bias will increase with surveys covering increasingly
larger redshift ranges, which is particularly important for LSST.}

@@

The relation between SNe~Ia and their host galaxy properties has been
studied extensively. The first key finding was that the standardized SNe~Ia
magnitudes significantly depend on the host galaxy stellar mass, with
SNe~Ia from high-mass host galaxies being brighter on average
\cite[e.g.][]{kelly2010, sullivan2010, childress2013, betoule2014, rigault2020,
kim19}. This mass-step correction is currently used in cosmological analyses
\citep[e.g.][]{betoule2014, scolnic2018a}, including for deriving $H_0$
\citep{riess2016, riess2019}. Yet, the underlying connection between the SNe and
their host galaxies remains , particularlyunclear when using
global properties such as the host stellar . This in turnmass raises
the question of the accuracy of such corrections since host galaxy
global properties evolve with redshift. More recently, studies have used the
local SN host galaxy environment to probe more direct connections
between the SNe and their host galaxy environments \citep{rigault2013},
showing that local age tracers such as the Local specific Star Formation Rate
(LsSFR) or the local color are more strongly correlated with the standardized SN
magnitude \citep{roman2018, kim18,rigault2020}. Further reinforcing this
connection is the identification of SN~Ia spectral features that are correlated
with LsSFR \citep{nordin2018}. These results suggest age as the driving
parameter underlying the mass-step. If true, this would have a significant
impact for cosmology, since the environmental correction to apply to SN
standardization could strongly vary with redshift. \citep{rigault2013,
childress2014, scolnic2018a}. Yet, the importance of local SN environmental
studies remains highly debated \cite[e.g.][]{jones2015, jones2019} and
especially the impact of such an astrophysical bias has on the derivation of
$H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the younger
(those with a high LsSFR) and which are the older (those with low LsSFR).
Furthermore, since the first SNe~Ia host analyses, the SN stretch has been known
to be strongly correlated with the SN host galaxy properties
\citep{hamuy1996, hamuy2000}. This correlation that has been extensively
confirmed since \citep[e.g.][]{neill2009, sullivan2010, lampeitl2010, kelly2010,
gupta2011, dandrea2011, childress2013, rigault2013, pan2014, kim19}. Following
the ``A+B'' model and the connection between SN stretch and host galaxy
properties, \cite{howell2007} first discussed the potential redshift drift of
the SN stretch distribution. In this paper we revisit this question
using the most recent SNe~Ia datasets.

In this paper, we take a step aside from the cosmological analyses to
probe the validity of our modeling of the SN population, which we claim to be
constituted of two age-populations \citep{rigault2013,rigault2015,rigault2020}:
one old and one younger, the former having on average lower lightcurve stretches
and being brighter after standardization. We use the correlation between the SN
age, as probed by the LsSFR, and the SN stretch to model the expected evolution
of the underlying SN stretch distribution as a function of redshift. This
modeling relies on three assumptions: (1) there are two distinct populations of
SNe~Ia; (2) the relative fraction of each of these populations as a function of
redshift follows the model presented in \cite{rigault2020} and (3) the
underlying distribution of stretch for each age sample is constant. This paper
aims at testing this specific model with datasets from the literature.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/zmax_maglim_snls.pdf}
    \caption{\textsc{\texttt{SALT2.4}} stretch ($x_1$) and color ($c$)
        lightcurve parameters of SNe~Ia from the SDSS, PS1 and SNLS samples of
        the Pantheon catalog. The individual SNe are shown as blue dots. The
        ellipse $(x_1=\pm3, c=\pm0.3)$ is displayed, colored by the
        corresponding standardized absolute magnitude using the $\alpha$ and
        $\beta$ coefficients from \cite{scolnic2018a}. The grey diagonal lines
        represent the $(x_1, c)$ evolution for $m = m_{\lim}$, for $z$ between
        $0.50$ and $z=1.70$ using SNLS's $m_{\lim}$ of $24.8$~mag.}
    \label{fig:maglim}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from the
        SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}). The colored parts represent the distribution of
        SNe~Ia kept in our analysis for they are supposedly free from
    observational selection bias (see Section~\ref{sec:sample}). The
darker (resp. lighter) color responds to the conservative (resp. fiducial)
selection cut.}
    \label{fig:cuts}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp. bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions}
    \label{fig:distrib}
\end{figure}

@@

\cite{rigault2020} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia follows
the number of Gyr-old stars in the Universe, i.e. the stellar mass (M$^*$).
Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the fraction of
young (resp. old) SNe~Ia in the Universe as a function of redshift, then the
ratio $\delta/\psi$ is expected to follow the evolution of the specific star
formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013, rigault2020, wiseman2020}, in agreement with rate
expectations \citep{mannucci2006, rodney2014}, \cite{rigault2020} concluded that

@@

Results of fitting this model to all the 569 (resp. 422) SNe from the fiducial
(resp. conservative) sample are given Table~\ref{tab:modelresults}, and the
predicted redshift evolution of mean stretch (expected $x_1$ given the
distribution of Eq.~\ref{eq:stretchz}) illustrated as a blue band in
Fig.~\ref{fig:modelall} accounting for parameters errors and their covariances.
We see in this figure that the measured mean SN~Ia stretch per redshift bins of
equal sample size closely follows our redshift drift modeling. This is indeed
what is expected if old environments favor low SN stretches
\citep[e.g.][]{howell2007} \textit{and} if the fraction of old SNe~Ia declines as
a function of redshift. See Section~\ref{sec:results} for a more quantitative
discussion.

\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{Article_figures/stretchevol_all_vs_snf-mean.pdf}
    \caption{Evolution of the mean SN \textsc{\texttt{SALT2.4}} stretch ($x_1$)
        as a function of redshift. Markers show the stretch plain mean
        (with error estimated from scatter) measured in redshift bins of equal
        sample size, indicated in light gray at the bottom of each redshift bin.
        Full and light markers are used when considering the fiducial or the
        conservative samples, respectively. The orange horizontal line
        represents the mean stretch for the non-evolving Gaussian model
        (last line of Table~\ref{tab:comp}) on the fiducial sample. Best
        fits of our Base drifting model are shown as blue, dashed-blue and
        gray, when fitted on the fiducial sample, the conservative one or
        the SNfactory dataset only, respectively; all are compatible. The
        light-blue band illustrates the amplitude of the error (incl.
        covariance) of the best fit model when considering the fiducial
    dataset.}
    \label{fig:modelall}
\end{figure*}

@@

In Section~\ref{sec:basemodel}, we have modeled the underlying stretch
distribution following \cite{rigault2020}, i.e.\ as a single Gaussian for the
``young'' SNe~Ia and a mixture of two Gaussians for the ``old'' SNe Ia
population, one being the same as for the young population, plus another one for
the fast-declining SNe~Ia that seem to only exist in old local environments.
This is our so-called ``Base'' model. However, to test different modeling
choices, we have implemented a suite of alternative parametrizations that we
also adjust to the data following the procedure described in
Section~\ref{sec:modelnopy}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Article_figures/bbc_comp_PS1_hist-nr.pdf}
    \caption{Distribution of the PS1 SN~Ia \textsc{\texttt{SALT2.4}} stretch
        ($x_1$) after the fiducial redshift limit cut (grey histogram). This
        distribution is supposed to be a random draw from the underlying stretch
        distribution. The green lines show the BBC model of this underlying
        distribution (asymmetric Gaussian). The full line (band) is our best fit
        (its error); the dashed line shows the \cite{scolnic2018a} result. The
        black line (band) shows our best fitted Base-modeling (its error, see
        Table \ref{tab:modelresults}) that includes redshift drift. For
        illustration, we show as colored (from blue to red with increasing
        redshifts) the evolution of the underlying stretch distribution as a
        function of redshift for the redshift range covered by PS1 data.}
    \label{fig:bbc_pdf_ps1}
\end{figure}

>>>>

\abstract{The detailed nature of type Ia supernovae (SNe~Ia) remains uncertain,
    and as survey statistics increase, the question of astrophysical systematic
    uncertainties rises, notably that of SN~Ia population evolution. In this
    paper, we study the dependence with redshift of the SN~Ia \texttt{SALT2.4}
    lightcurve stretch, a purely intrinsic SN property, to probe its potential
    redshift drift. The SN stretch has been shown to strongly correlate with the
    SN environment, notably with stellar age tracers. We model the underlying
    stretch distribution as a function of redshift, using the evolution of the
    fraction of young and old SNe~Ia as predicted \textbf{using the SNfactory
    dataset}, and assuming constant underlying stretch distribution for each age
    population consisting of Gaussian mixtures. We test our prediction against
    published samples cut to have marginal magnitude selection effects, so that
    any observed change is indeed astrophysical and not observational in origin.
    In this first study, there are indications that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a better description of the data than any time-constant
    model, including the sample-based asymmetric distributions often used to
    correct Malmquist bias at a significance of over 5 $\sigma$.  The favored
    underlying stretch model is \textbf{a bimodal one}, composed of a
    high-stretch mode shared by both young and old environments, and a
    low-stretch mode exclusive to old environments. The precise impact of the
    redshift evolution of the SN Ia population intrinsic properties on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
    distances derived from SN affected by observational selection effects. We
highlight that such a bias will increase with surveys covering increasingly
larger redshift ranges, which is particularly important for LSST.}

@@

The relation between SNe~Ia and their host galaxy properties has been studied
extensively. The first key finding was that the standardized SNe~Ia magnitudes
significantly depend on the host galaxy stellar mass, with SNe~Ia from high-mass
host galaxies being brighter on average \cite[e.g.][]{kelly2010, sullivan2010,
childress2013, betoule2014, kim19}. This mass-step correction is currently used
in cosmological analyses \citep[e.g.][]{betoule2014, scolnic2018a}, including
for deriving $H_0$ \citep{riess2016, riess2019}. Yet, the underlying connection
between the SNe and their host galaxies remains unclear, particularly when using
global properties such as the host stellar mass. This in turn raises the
question of the accuracy of such corrections since host galaxy global properties
evolve with redshift. More recently, studies have used the local SN host galaxy
environment to probe more direct connections between the SNe and their host
galaxy environments \citep{rigault2013}, showing that local age tracers such as
the Local specific Star Formation Rate (LsSFR) or the local color are more
strongly correlated with the standardized SN magnitude \citep{roman2018, kim18,
rigault2020}. Further reinforcing this connection is the identification of SN~Ia
spectral features that are correlated with LsSFR \citep{nordin2018}. These
results suggest age as the driving parameter underlying the mass-step. If true,
this would have a significant impact for cosmology, since the environmental
correction to apply to SN standardization could strongly vary with redshift
\citep{rigault2013, childress2014, scolnic2018a}. Yet, the importance of local
SN environmental studies remains highly debated \cite[e.g.][]{jones2015,
jones2019} and especially the impact of such an astrophysical bias has on the
derivation of $H_0$ \citep{jones2015, riess2016, riess2018, rose2019}.

The concept of the SN~Ia age dichotomy arose with the study of the SN~Ia rate.
\cite{mannucci2005, scannapieco2005, sullivan2006, aubourg2008} have shown that
the relative SNe~Ia rate in galaxies could be explained if two populations
existed, one young, following the host star formation activity, and one old
following the host stellar mass (the so called ``prompt and delayed'' or ``A+B''
model). \cite{rigault2020} used the LsSFR to classify which are the younger
(those with a high LsSFR) and which are the older (those with low LsSFR).
Furthermore, since the first SNe~Ia host analyses, the SN stretch has been known
to be strongly correlated with the SN host galaxy properties \citep{hamuy1996,
hamuy2000}. This correlation that has been extensively confirmed since
\citep[e.g.][]{neill2009, lampeitl2010, gupta2011, dandrea2011, pan2014}.
Following the ``A+B'' model and the connection between SN stretch and host
galaxy properties, \cite{howell2007} first discussed the potential redshift
drift of the SN stretch distribution. In this paper we revisit this question
using the most recent SNe~Ia datasets.

In this paper, we take a step aside from the cosmological analyses to probe the
validity of our modeling of the SN population, which we claim to be constituted
of two age-populations \citep{rigault2013, rigault2015, rigault2020}: one old
and one younger, the former having on average lower lightcurve stretches and
being brighter after standardization. We use the correlation between the SN age,
as probed by the LsSFR, and the SN stretch to model the expected evolution of
the underlying SN stretch distribution as a function of redshift. This modeling
relies on three assumptions: (1) there are two distinct populations of SNe~Ia;
(2) the relative fraction of each of these populations as a function of redshift
follows the model presented in \cite{rigault2020} and (3) the underlying
distribution of stretch for each age sample is constant. This paper aims at
testing this specific model with datasets from the literature.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/zmax_maglim_snls.pdf}
    \caption{\textsc{\texttt{SALT2.4}} stretch ($x_1$) and color ($c$)
        lightcurve parameters of SNe~Ia from the SDSS, PS1 and SNLS samples of
        the Pantheon catalog. The individual SNe are shown as blue dots. The
        ellipse $(x_1=\pm3, c=\pm0.3)$ is displayed, colored by the
        corresponding standardized absolute magnitude using the $\alpha$ and
        $\beta$ coefficients from \cite{scolnic2018a}. The gray diagonal lines
        represent the $(x_1, c)$ evolution for $m = m_{\lim}$, for $z$ between
        $0.50$ and $z=1.70$ using SNLS's $m_{\lim}$ of $24.8$~mag.}
    \label{fig:maglim}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/hist_surveys_cuts_55-cividis.pdf}
    \caption{\textit{From top to bottom}: Redshift histograms of SNe~Ia from the
        SDSS, PS1 and SNLS dataset respectively (data from Pantheon,
        \citealt{scolnic2018a}). The colored parts represent the distribution of
        SNe~Ia kept in our analysis for they are supposedly free from
    observational selection bias (see Section~\ref{sec:sample}). The
darker (resp.\ lighter) color responds to the conservative (resp.\ fiducial)
selection cut.}
    \label{fig:cuts}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp.\ bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions}
    \label{fig:distrib}
\end{figure}

@@

\cite{rigault2020} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia follows
the number of Gyr-old stars in the Universe, i.e. the stellar mass (M$^*$).
Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the fraction of
young (resp.\ old) SNe~Ia in the Universe as a function of redshift, then the
ratio $\delta/\psi$ is expected to follow the evolution of the specific star
formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013, rigault2020, wiseman2020}, in agreement with rate
expectations \citep{mannucci2006, rodney2014}, \cite{rigault2020} concluded that

@@

Results of fitting this model to all the 569 (resp. 422) SNe from the fiducial
(resp.\ conservative) sample are given Table~\ref{tab:modelresults}, and the
predicted redshift evolution of mean stretch (expected $x_1$ given the
distribution of Eq.~\ref{eq:stretchz}) illustrated as a blue band in
Fig.~\ref{fig:modelall} accounting for parameters errors and their covariances.
We see in this figure that the measured mean SN~Ia stretch per redshift bins of
equal sample size closely follows our redshift drift modeling. This is indeed
what is expected if old environments favor low SN stretches
\citep[e.g.][]{howell2007} \textit{and} if the fraction of old SNe~Ia declines as
a function of redshift. See Section~\ref{sec:results} for a more quantitative
discussion.

\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{Article_figures/stretchevol_all_vs_snf-mean.pdf}
    \caption{Evolution of the mean SN \textsc{\texttt{SALT2.4}} stretch ($x_1$)
        as a function of redshift. Markers show the stretch plain mean
        (with error estimated from scatter) measured in redshift bins of equal
        sample size, indicated in light gray at the bottom of each redshift bin.
        Full and light markers are used when considering the fiducial or the
        conservative samples, respectively. The orange horizontal line
        represents the mean stretch for the non-evolving Gaussian model
        (last line of Table~\ref{tab:comp}) on the fiducial sample. Best
        fits of our Base drifting model are shown as blue, dashed-blue and
        gray, when fitted on the fiducial sample, the conservative one or
        the SNfactory dataset only, respectively; all are compatible. The
        light-blue band illustrates the amplitude of the error (incl.\
        covariance) of the best fit model when considering the fiducial
    dataset.}
    \label{fig:modelall}
\end{figure*}

@@

In Section~\ref{sec:basemodel}, we have modeled the underlying stretch
distribution following \cite{rigault2020}, i.e.\ as a single Gaussian for the
``young'' SNe~Ia and a mixture of two Gaussians for the ``old'' SNe Ia
population, one being the same as for the young population, plus another one for
the fast-declining SNe~Ia that seem to only exist in old local environments.
This is our so-called ``Base'' model. However, to test different modeling
choices, we have implemented a suite of alternative parameterizations that we
also adjust to the data following the procedure described in
Section~\ref{sec:modelnopy}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Article_figures/bbc_comp_PS1_hist-nr.pdf}
    \caption{Distribution of the PS1 SN~Ia \textsc{\texttt{SALT2.4}} stretch
        ($x_1$) after the fiducial redshift limit cut (gray histogram). This
        distribution is supposed to be a random draw from the underlying stretch
        distribution. The green lines show the BBC model of this underlying
        distribution (asymmetric Gaussian). The full line (band) is our best fit
        (its error); the dashed line shows the \cite{scolnic2018a} result. The
        black line (band) shows our best fitted Base-modeling (its error, see
        Table \ref{tab:modelresults}) that includes redshift drift. For
        illustration, we show as colored (from blue to red with increasing
        redshifts) the evolution of the underlying stretch distribution as a
        function of redshift for the redshift range covered by PS1 data.}
    \label{fig:bbc_pdf_ps1}
\end{figure}
